Comparison of large scale citizen science data and long term study data for phenology modeling

### 2018-01-17 manuscript/main.tex
Spatial model analysis (#35)
-add spatial models (MSB and M1) to figures
-add map showing all the sites
-update manuscript intro structure to ethans suggestions

%% \title{Testing the usability of large scale citizen science data for phenology modeling}
\title{Comparison of large scale citizen science data and long term study data for phenology modeling}
\author[1]{Shawn D. Taylor}
\author[2]{Ethan P. White}
\affil[1]{UFL}
\affil[2]{UFL}
\corrauthor[1]{Shawn D. Taylor}{shawntaylor@weecology.org}

@@

Citizen science data is becoming increasingly important for ecological research, with the potential for significant impact in many fields \citep{dickinson2010, tulloch2013, kelling2009}. Citizen science efforts have helped document range shifts \citep{hitch2007}, study the effect of landscape fragmentation on amphibians \citep{cosentino2014}, and informed ecological theory of community composition \citep{locey2013}. A primary benefit of citizen science data is observations made at numerous sites across regional to continental scales. This helps researchers explain large scale patterns across natural gradients, but can also introduce new sources error because of observer skill, variation in sampling effort, and spatial bias \citep{dickinson2010}. These issues can be accounted for in models but first must be identified and quantified. As new citizen science datasets are made available, they must be evaluated for their own biases and limitations.

The National Phenology Network (NPN) is an organization which accepts observations of phenology from volunteers throughout North America and makes the data freely available \citep{schwartz2012a}. Plant phenology is important for studies of ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}. Robust phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant and animal communities. At the largest scale variation in the timing of spring leaf out and fall senescence account for a significant amount of error in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long term forecasts \citep{richardson2012}. At local scales, because of species specific responses to temperature and precipitation, changes in phenology can alter the flower community available to pollinators and seed predators throughout the growing season \citep{diez2012, caradonna2014, ogilvie2017, theobald2017}.

The NPN dataset has already being used to study, among many things, variation in oak phenology \citep{gerst2017}, large scale community phenology models \citep{melaas2016}, and forecasting long term phenology trends \citep{jeong2013}. While there has been several studies looking at best practices when using this dataset \citep{crimmins2017, gerst2016} no study to date has looked a the underlying validity of the data. With hundreds of observers across North America, the observer and sampling bias for species identification and the true date of phenological events is potentially high. Though volunteers can be very accurate at distinguishing different phenophases \citep{fuccillo2015} observations can still be made sporadically throughout the growing season, some years skipped, or visits to a particular site discontinued entirely. These biases could adversely affect process based phenology models, which are species specific and depend on high precision in the data \citep{chuine2000}.

The NPN dataset is complicated further by the spatial variation of the temperature response within species. Phenology models currently have trouble capturing the spatial variation in warming requirements \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}. Most phenology models are process based and designed to reflect physological processes driving phenological events. Yet models parameterized using observations from one site rarely transfer to other sites, and models using pooled data from many sites do not transfer well to individual sites \citep{garcia-mozo2008, basler2016}. This variation within species can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995,diez2012}. While this complicates further any testing of dataset validity, it's worth it because there is a large opportunity with this dataset for further study of phenology across spatial gradients.

Many phenological studies in the past have relied upon datasets with observations collected at a single location over a long timeframe by a single group \citep{cook2012, roberts2015, clark2014}. These datasets are assumed to not suffer from the potential biases mentioned above, and many also have high temporal coverage of several decades. They are disadvantageous as the inference from the models can only be applied to species located at the site, and only to local climate variability. These datasets have been invaluable for ecology for several decades, but datasets with a larger spatial extent and more complete species pool will be needed to model the complex phenology dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}. The NPN datasets has the potential to meet these needs, and to date no study has assessed how well NPN data compares to long term study phenology datasets. Here we will compare process based phenology models built using only the NPN data with models built with data from Long Term Study (LTS) datasets. The specific research questions are:
\begin{enumerate}
\item Do parameter estimates of species level models differ when they are built with the NPN dataset versus an LTS dataset?
\item Do estimates of phenological events differ between the NPN and LTS derived models?
\item How well do models built with the NPN dataset transfer to LTS observations, and vice verse?
\end{enumerate}

@@

To compare against the NPN dataset we used four LTS datasets from North America (Table 1). Three major ecosystem types and 24 species are represented in these datasets. As all four have been collected by single groups we make the assumption that they are free of any bias in species identification or sampling effort. Observation metrics varied among these four datasets due to different protocols, so were first converted to binary "yes" and "no" observations for each phenophase (see supplement for details). As with the NPN dataset we infer the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation.

@@

The naive model uses only the mean DOY from all prior observations as the estimated DOY. The linear model uses a regression with the mean spring (Jan. - March) temperature as the only independent variable and DOY as the response variable.

\begin{center}
{\def\arraystretch{2}\tabcolsep=5pt
    \begin{tabular}{ | l | c | c | p{1.3cm} | l |}
    \hline
    Name & DOY Estimator & Forcing Equations & Total\newline Parameters & Reference \\ \hline
    Naive & \( \overline{DOY} \) & - & 1 & - \\
    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(x_{t})\geq F* $  & $R_{f}(x_{t}) = min(T, 0)$ & 1 & - \\
    Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(x_{t})\geq F* $ & $ R_{f}(x_{t}) = max(T - T^{*}, 0) $  & 3 & - \\
    Alternating & $\sum_{t=0}^{DOY}R_{f}(x_{t})\geq a + be^{cNCD(t)} $ & $R_{f}(x_{t}) = min(T, 0) $ & 3 & \citep{cannell1983} \\
    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(x_{t})\geq F* $ & $ R_{f}(x_{t}) = \frac{1}{1 + e^{b(x-c)}} $ & 4 & \citep{chuine2000} \\

@@

Parameterization was done using differential evolution to minimize the root mean square error (RMSE) of the estimated day of year of the phenological event. Differential evolution is a global optimization algorithm similar to simulated annealing, which has traditionally been used to fit complex phenology models \citep{storn1997, chuine2000}. Initial testing showed that differential evolution produced more consistent results than simulated annealing and other similar algorithms. Confidence intervals for parameters were obtained by bootstrapping, where individual models were re-fit 250 times using a random sample, with replacement, of the data. A random 20\% of all data was held out from model fitting for use in later evaluation.

@@

As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set parameterized using only LTS data. The distribution for each parameter in each model was then compared between the two sets using a Mann-Whitney test. This was chosen due to the majority of parameters having non-normal distributions derived from the bootstrapped estimates. Statistically significant differences in these parameters, which are designed to represent biological mechanisms, would imply that the LTS and NPN datasets do not agree on

Even models with different structures can produce similar estimates of phenological events \citep{basler2016}. Thus we also compared estimates between LTS and NPN derived models using the root mean square difference (RMSD) between the two estimates:

@@

where $n$ is the number of estimates produced for a single species and phenophase, and $LTS_{i}$ and $NPN_{i}$ are DOY estimates produced from the respective models. A RMSD value close to 0 implies that the LTS and NP derived models produced very similar estimates. For each species and phenophase 12 RMSD values were calculated from the 6 phenology models as well as 2 spatial extents. The first spatial extent used estimates from a the localities of the respective species in the NPN dataset, thus representing a large spatial extent. The second used estimates from the localities of the LTS datasets, thus representing a single site. The RMSD does not consider actual observations as it is meant to infer the correspondence of the NPN and LTS datasets beyond parameter comparison.

For all models the actual error was calculated using the using the RMSE:

@@

For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, held out NPN and LTS observations, and the six phenology models.

Pairwise model error comparison

@@

@@

-Even with different parameter values, some species/models had extremely similar
estimates, suggesting these models had

@@

Even with similar parameters, the Fixed GDD model was not the best performing model when tested on out of sample data. While the values of additional parameters may be incorrect in the above sense, they nonetheless provide more predictive power. Across the two spatial scales the GDD, Uniforc, and Alternating model were generally the best performing (Fig. 3). Considering only NPN derived models predicting out of sample NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1). One goal of ecological forecasting is to inform decision makers about the future state of systems (dietz?). Thus the best performing model should be used in these cases to give the best possible forecast. The GDD and Uniforc models have three and four parameters, respectively. Thus the simpler GDD model shows promise for building models with the NPN dataset and making forecasts across large scales.

@@

The spatially corrected NPN models were essentially identical to the non-corrected ones in terms of parameters, estimates of phenological events, and out of sample error rates. Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, yet models which attempt to model this variation did not significatnly improve estimates

\section*{Conclusion}
Despite being collected across a large spatial extend and by hundreds of observers the NPN dataset can be used to build phenology models which make comparable predictions to LTS datasets. Indeed, if making predictions across a large spatial extent the NPN derived models are preferable to LTS derived models. If using the NPN dataset to make inferences about plant physiology researchers should not to use overly complicated models, as parameter estimates from complex NPN derived models do not agree with LTS derived models. Whether this disagreement stems from the spatial variation in plant phenology requirements, or in the irregular sampling of the NPN dataset is an avenue for future study.

The National Phenology Network dataset will likely be an important resource for studying phenology.  The addition of numerous species and observations from across a range of conditions is important for informing large scale processes, as similar datasets have shown in Europe \citep{olsson2014, basler2016} and Asia \citep{xu2013, zhang2017}. Research from long term small scale studies has been invaluable and should no doubt be continued, but can now be complimented by using large scale NPN data (ie. \citep{jeong2013, melaas2016}). As long term phenology observations from the National Ecological Observatory Network \citep{elmendorf2016} become available in the coming years  ....

@@

>>>>

\title{Comparison of large scale citizen science data and long term study data for phenology modeling}

\author[1]{Shawn D. Taylor}
\author[2]{Ethan P. White}
\affil[1]{UFL}
\affil[2]{UFL}
\corrauthor[1]{Shawn D. Taylor}{shawntaylor@weecology.org}

@@

Plant phenology affects many fields of ecological research such as ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}. Robust phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant and animal communities. At the largest scale variation in the timing of spring leaf out and fall senescence account for a significant amount of error in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long term forecasts \citep{richardson2012}. At local scales, because of species specific responses to temperature and precipitation, changes in phenology can alter the flower community available to pollinators and seed predators throughout the growing season, potentially causing a cascade of affects on abundance, species richness, and interspecific interactions. \citep{diez2012, caradonna2014, ogilvie2017, theobald2017}.

Many phenological studies in the past have relied upon datasets collected at a single location over a long time-frame by a single group \citep{cook2012, roberts2015, iler2013, wolkovich2012}. These datasets have high temporal coverage, with some approaching several decades, and represent the species and climate variability at their respective sites extremely well. Yet it is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}. Phenology models are species specific, thus this lack of transferability across space can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995,diez2012}. Models fit to only a single site can easily be overfit, and being able to generalize across a species range will be important for improving phenology forecasts \citep{richardson2013}. To accomplish this phenology observations will be needed from as much of the full range of species as possible.

The National Phenology Network (NPN) is an organization which accepts phenology observations from volunteers throughout North America and makes the data freely available \citep{schwartz2012a}. Citizen science data such as this is becoming increasingly important for ecological research, with the potential for significant impact in many fields \citep{dickinson2010, tulloch2013, kelling2009}. Citizen science efforts have helped document range shifts \citep{hitch2007}, study the effect of landscape fragmentation on amphibians \citep{cosentino2014}, and informed ecological theory of community composition \citep{locey2013}. The NPN dataset has already being used to study, among many things, variation in oak phenology \citep{gerst2017}, large scale community phenology models \citep{melaas2016}, and forecasting long term phenology trends \citep{jeong2013}. A primary benefit of citizen science data is observations made at numerous sites across regional to continental scales. This helps researchers explain large scale patterns across natural gradients, but can also introduce new sources error from observer skill, variation in sampling effort, and spatial bias \citep{dickinson2010}. These issues can be accounted for in models but first must be identified and quantified. While there has been several studies looking at best practices when using the NPN dataset \citep{crimmins2017, gerst2016} no study to date has looked a the underlying validity of the data. With hundreds of observers across North America, the observer and sampling bias for species identification and the true date of phenological events is potentially high. Though volunteers can be very accurate at distinguishing different phenophases \citep{fuccillo2015}, observations can still be made sporadically throughout the growing season, some years skipped, or visits to a particular site discontinued entirely.

Data from long term studies have been invaluable for phenology research for several decades, but datasets with a larger spatial extent and more complete species pool will be needed to model the complex phenology dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}. Such datasets from China and Europe have already contributed considerably to phenological research \citep{olsson2014, basler2016, xu2013, zhang2017}, and the NPN dataset has the potential to meet these needs for North America. To date no study has assessed how well the NPN dataset compares to long term study phenology datasets. Here we will compare process based phenology models built using only the NPN data with models built with data from Long Term Study (LTS) datasets. The specific research questions are:
\begin{enumerate}
\item Do parameter estimates of species level models differ when they are built with the NPN dataset versus an LTS dataset?
\item Do estimates of phenological events differ between the NPN and LTS derived models?
\item How well do models built with the NPN dataset transfer to LTS observations, and vice verse?
\end{enumerate}

@@

To compare against the NPN dataset we used four LTS datasets from North America (Table 1). Three major ecosystem types and 24 species are represented in these datasets. Observation metrics varied among these four datasets due to different protocols, so were first converted to binary "yes" and "no" observations for each phenophase (see supplement for details). As with the NPN dataset we infer the date for each phenophase as the midpoint between the first "yes" observation and most recent "no" observation.

@@

The naive model uses only the mean DOY from all prior observations as the estimated DOY. The linear model uses a regression with the mean spring (Jan. - March) temperature as the only independent variable and DOY as the response variable. Finally two models attempt to explain the spatial variation in phenological requirements. The first spatial model, M1, is an extension of the GDD model which adds a correction in the required forcing using the photoperiod ($L$). The second, the Macroscale Species-specific Budburst model (MSB), uses the mean spring temperature as a linear correction on the total forcing required in the Alternating model. Since there is little to no spatial variation in the LTS datasets, the two spatial models were fit using only the NPN dataset. The resulting parameters, estimates, and errors for the NPN derived M1 and MSB models were compared to the LTS GDD and Alternating models, respectively, for each species and phenophase.

\begin{center}
{\def\arraystretch{2}\tabcolsep=5pt
    \begin{tabular}{ | l | c | c | p{1.3cm} | l |}
    \hline
    Name & DOY Estimator & Forcing Equations & Total\newline Parameters & Reference \\ \hline
    Naive & \( \overline{DOY} \) & - & 1 & - \\
    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = min(T_{i}, 0)$ & 1 & - \\
    Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T^{*}, 0) $  & 3 & - \\
    M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T^{*}, 5) $  & 4 & \citep{blumel2012} \\
    Alternating & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD(t)} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 3 & \citep{cannell1983} \\
    MSB & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD_{i}} +dT_{mean} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 4 & \citep{jeong2013} \\
    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = \frac{1}{1 + e^{b(T_{i}-c)}} $ & 4 & \citep{chuine2000} \\

@@

All models were parameterized using differential evolution to minimize the root mean square error (RMSE) of the estimated day of year of the phenological event. Differential evolution is a global optimization algorithm similar to simulated annealing, which has traditionally been used to fit complex phenology models \citep{storn1997, chuine2000}. Initial testing showed that differential evolution produced more consistent results than simulated annealing and other similar algorithms. Confidence intervals for parameters were obtained by bootstrapping, where individual models were re-fit 250 times using a random sample, with replacement, of the data. A random 20\% of all data was held out from model fitting for use in later evaluation. Predictions were made by taking the mean DOY from the 250 bootstrapped models.

@@

As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set parameterized using only LTS data (with the exception of the M1 and MSB models, see above). The distribution for each parameter in each species/phenophase model was then compared between the two sets using a Mann-Whitney test. This was chosen due to the majority of parameters having non-normal distributions derived from the bootstrapped estimates.

Even models with entirely different structures can produce similar estimates of phenological events \citep{basler2016}. Thus we also compared estimates of the phenological events between LTS and NPN derived models using the root mean square difference (RMSD) between the two estimates:

@@

where $n$ is the number of estimates produced for a single species and phenophase, and $LTS_{i}$ and $NPN_{i}$ are DOY estimates produced from the respective models. A RMSD value close to 0 implies that the LTS and NPN derived models produced very similar estimates. For each species and phenophase 12 RMSD values were calculated from the 6 phenology models and 2 scales. We calculated the RMSD for estimates at the first scale of the extent of the NPN observations, thus representing a large spatial extent. The second scale used estimates from the localities of the LTS datasets, thus representing a single site. The RMSD does not consider actual observations as it is meant to infer the correspondence of the NPN and LTS datasets beyond parameter comparison.

For all models the observation error of the held out data was calculated using the using the RMSE:

@@

For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, the six phenology models, and the two scales described above.

TODO: Pairwise model error comparison

@@

TODO: figure 4 results

@@

@@

Even with similar parameters, the Fixed GDD model was not the best performing model when tested on out of sample data. While the values of additional parameters may be incorrect in the above sense, they nonetheless provide more predictive power. Across the two scales the GDD, Uniforc, and Alternating model were generally the best performing (Fig. 3). Considering only NPN derived models predicting NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1). One goal of ecological forecasting is to inform decision makers about the future state of systems \citep{dietz2017}. Thus the best performing model should be used in these cases to give the best possible forecast. The GDD and Uniforc models have three and four parameters, respectively, therefore the simpler GDD model shows promise for building models with the NPN dataset and making forecasts across large scales.

@@

The models which incorporate a spatial component did not improve over their base models. Correspondence between parameter estimates, estimates of phenological events, and out of sample error rates for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively. The MSB and M1 models had the lowest error in 9 out of 39 occasions, but never by more than a single day over other models (Fig S1). Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, yet here models which attempt to account for this variation did not improve estimates overall. This could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors. \cite{basler2016} used the M1 model to predict budburst on six species across Europe and found it was usually among the best models in terms of RMSE, albeit never by more than a single day. Their result was strengthened by a 40 year timeseries across a large region.

\section*{Conclusion}

Currently the best phenology model for a single location will always be those fit using data from only that location, yet these single site models do not transfer well to large scales. Making large scale phenology forecasts will require incorporating large scale observations such as those from the NPN dataset. Here we have shown that the NPN dataset is capable of parameterizing phenology models which make comparable predictions to LTS datasets. Indeed, if making forecasts across a large spatial extent the NPN derived models are preferable to LTS derived models. If using the NPN dataset to make inferences about plant physiology researchers should not use overly complicated models, as parameter estimates from complex NPN derived models do not agree with LTS derived models. Whether this disagreement stems from the spatial variation in plant phenology requirements, the irregular sampling of the NPN dataset, or the non-identifiability of models is an avenue for future study.

The National Phenology Network dataset will likely be an important resource for studying phenology.  The addition of numerous species and observations from across a range of conditions is important for informing large scale processes, as similar datasets have shown in Europe \citep{olsson2014, basler2016} and Asia \citep{xu2013, zhang2017}. Research from long term small scale studies has been invaluable and should no doubt be continued, but can now be complimented by using large scale NPN data (ie. \citep{jeong2013, melaas2016}).

@@

\setcounter{figure}{0}    
\section*{Suppliment}
\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{fig_s1_best_npn_models.png}
	\caption{Model error for specific species and phenophases of the NPN dataset. X marks the best performing models for the respective data type.}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{fig_s3_site_map.png}
	\caption{Locations of National Phenology Network sites used (black points) and Long Term Study sites (labeled circles).}
\end{figure}
